---
title: "Time series analysis: annual maximum streamflows"
output: html_document
---

# **Call libraries**

```{r callLib}
library(readr)
library(dplyr)
library(moments)   # provides skewness() and kurtosis()
library(reshape2)
library(forecast)
library(gridExtra)
library(knitr)
library(ggplot2)
library(tseries)
library(nortest)
library(kableExtra)
library(randtests)
library(trend)
```

# **Read data**

## Read zip file names

```{r rzip}
zip_files <- list.files("/home/alejandro/Documents/unal/lectures/stochasticMethodsInWR/scripts/data/ElBancoMagMaxAnnualDisch", pattern = "\\.zip$", full.names = TRUE)
```

## Read one CSV per zip

```{r rcsv}
data_list <- lapply(zip_files, function(zf) {
  csv_name <- unzip(zf, list = TRUE)$Name[1]   # take the first CSV in zip
  read_csv(unz(zf, csv_name))
})
```

## Combine into one dataframe

```{r cindf}
data_all <- bind_rows(data_list)
```

## Covert a char date column into date

```{r formatdata}
data_all$date <- as.Date(data_all$Fecha, format = "%d/%m/%Y")
data_all$year <-as.numeric(format(data_all$date, "%Y"))
```

# **Time series preprocessing**

## Display first rows

```{r displaydata}
head(data_all)
X = data_all$Valor
```

## Show the data structure

```{r datastr}
str(data_all)
```

## Compute statistics summary

```{r computestats}
summary_stats <- data.frame(
  Mean = mean(X, na.rm = TRUE),
  SD = sd(X, na.rm = TRUE),
  Min = min(X, na.rm = TRUE),
  Max = max(X, na.rm = TRUE),
  Median = median(X, na.rm = TRUE),
  Skewness = skewness(X, na.rm = TRUE),
  Kurtosis = kurtosis(X, na.rm = TRUE),
  N = length(X)
)
```

## Print stats

```{r printstats, echo=FALSE}

# Transpose the summary data frame
summary_vertical <- data.frame(
  Statistic = names(summary_stats),
  Value = as.numeric(summary_stats[1, ])
)
kable(summary_vertical, digits = 2, caption = "Summary Statistics of the Time Series")
```

## Plot time series

```{r plotts}
tit <- paste(unique(data_all$Parametro), " at ", unique(data_all$NombreEstacion), sep="") 
ggplot(data_all, aes(x = date, y = Valor)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point(color = "darkblue") +
  labs(title = tit, x = "Year", y = "Q (m^3 /s)") +
  theme_minimal()
```

## Plot histogram

```{r plothist}
# Calculate summary statistics
q <- quantile(data_all$Valor, probs = c(0.25, 0.5, 0.75))  # Q1, Median, Q3

# Compute density-based histogram data
h <- hist(X, plot = FALSE)
dens <- h$counts / sum(h$counts) / diff(h$breaks[1:2])  # convert counts to density
ymax <- max(dens)  # max histogram density

ggplot(data_all, aes(x = Valor)) +
  geom_histogram(aes(y = ..density..), bins = 15,
                 fill = "skyblue", color = "white", alpha = 0.6) +
  
  # Density overlay
  geom_density(color = "darkblue", size = 1.2, alpha = 0.7) +
  
  # Vertical lines for mean & quartiles
  geom_vline(xintercept = summary_stats$Mean, color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = q[1], color = "darkgreen", linetype = "dotted", size = 1) +
  geom_vline(xintercept = q[2], color = "blue", linetype = "solid", size = 1) +
  geom_vline(xintercept = q[3], color = "darkgreen", linetype = "dotted", size = 1) +
  
  # Set limits to Y axis
  coord_cartesian(ylim = c(0, ymax)) +
  
  # Labels for stats
  annotate("text", x = summary_stats$Mean, y = 0.02, label = "Mean", color = "red", angle = 90, vjust = -0.5) +
  annotate("text", x = q[2], y = 0.02, label = "Median", color = "blue", angle = 90, vjust = -0.5) +
  
  labs(
    title = "Streamflow Histogram + Density with Mean & Quartiles",
    x = "Q (m³/s)",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)
```

## Q–Q Plot of Time Series

```{r qqplot, echo=TRUE, fig.width=6, fig.height=5}
ggplot(data_all, aes(sample = Valor)) +
  stat_qq(color = "steelblue", size = 2) +
  stat_qq_line(color = "red", linetype = "dashed") +
  labs(title = "Normal Q-Q Plot",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

## Normality Summary of the Time Series

```{r normality-summary, message=FALSE, warning=FALSE}
# Normality tests
sw <- shapiro.test(X)
ad <- ad.test(X)
jb <- jarque.bera.test(X)

# Combine results
results <- data.frame(
  Statistic = c("Mean", "Std. Deviation", "Skewness", "Kurtosis",
                "Shapiro-Wilk (p)", "Anderson-Darling (p)", "Jarque-Bera (p)"),
  Value = c(
    summary_stats$Mean,
    summary_stats$SD,
    summary_stats$Skewness,
    summary_stats$Kurtosis,
    sw$p.value,
    ad$p.value,
    jb$p.value
  ),
  Interpretation = c(
    "", "", 
    ifelse(abs(summary_stats$Skewness) < 0.5, "≈ Symmetric", 
           ifelse(summary_stats$Skewness > 0, "Right-skewed", "Left-skewed")),
    ifelse(abs(summary_stats$Kurtosis - 3) < 0.5, "≈ Normal tails", 
           ifelse(summary_stats$Kurtosis > 3, "Leptokurtic (heavy tails)", "Platykurtic (light tails)")),
    ifelse(sw$p.value > 0.05, "Normal", "Not normal"),
    ifelse(ad$p.value > 0.05, "Normal", "Not normal"),
    ifelse(jb$p.value > 0.05, "Normal", "Not normal")
  )
)

kable(results, digits = 4, caption = "Summary statistics and normality tests for the time series")
```

## Normalization of the Time Series

```{r normalize-ts}
# Example: standardize time series
Xn <- scale(X)
data_all$ValorN <- Xn

# Check new mean and sd
mean(Xn, na.rm = TRUE)
sd(Xn, na.rm = TRUE)

```

## Comparison of Original and Normalized Time Series

```{r normalized-plot, message=FALSE, warning=FALSE}

# Prepare data
df <- data.frame(
  time = data_all$date,
  Original = as.numeric(X),
  Normalized = as.numeric(Xn)
)

# Plot with dual y-axes
ggplot(df, aes(x = time)) +
  geom_line(aes(y = Original, color = "Original"), size = 1) +
  geom_line(aes(y = Normalized * sd(X) + mean(X), color = "Normalized"), size = 1) +
  scale_y_continuous(
    name = "Original values",
    sec.axis = sec_axis(~ (. - mean(X)) / sd(X),
                        name = "Normalized (Z-score)")
  ) +
  scale_color_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Original and Normalized Time Series",
    x = "Year",
    color = "Series"
  ) +
  theme_minimal() +
  theme(
    axis.title.y.left = element_text(color = "steelblue"),
    axis.title.y.right = element_text(color = "tomato"),
    legend.position = "top"
  )
```

## Q–Q Plot of Normalized Time Series

```{r qqplot_normalized, echo=TRUE, fig.width=6, fig.height=5}
ggplot(data_all, aes(sample = ValorN)) +
  stat_qq(color = "steelblue", size = 2) +
  stat_qq_line(color = "red", linetype = "dashed") +
  labs(title = "Normal Q-Q Plot",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

# Time series analysis

## Randomness Tests

```{r randomness-tests, message=FALSE, warning=FALSE}


# Apply tests
runs <- runs.test(X)
tp   <- turning.point.test(X)
lb   <- Box.test(X, lag = 20, type = "Ljung-Box")

# Combine results
results <- data.frame(
  Test = c("Runs Test", "Turning Point Test", "Ljung-Box (lag=20)"),
  Statistic = c(runs$statistic, tp$statistic, lb$statistic),
  p_value = c(runs$p.value, tp$p.value, lb$p.value),
  Decision = ifelse(c(runs$p.value, tp$p.value, lb$p.value) > 0.05,
                    "Random (fail to reject H₀)",
                    "Not random (reject H₀)")
)

knitr::kable(results, digits = 4, caption = "Tests for randomness in the time series")
```

## Trend Analysis for Annual Time Series

```{r ts-annual-trend, message=FALSE, warning=FALSE}
# Convert to time series
x <- ts(data_all$Valor, start = min(data_all$year), frequency = 1)

# --- Trend estimation using a moving average ---
trend <- stats::filter(x, rep(1/3, 3), sides = 2)

# Plot
plot(x, col = "steelblue", lwd = 2, main = "Annual Time Series with Trend", ylab = "Q (m³/s)", xlab = "Year")
lines(trend, col = "tomato", lwd = 2)
legend("topleft", legend = c("Original", "3-Year Moving Average"), col = c("steelblue", "tomato"), lwd = 2)
```

## Detrending an Annual Time Series and Testing for Trend

In this section, we detrend an annual time series by fitting a linear model, visualize the fitted trend, and test for a significant monotonic trend using the Mann–Kendall test.

```{r detrend-trendtest-summary, echo=TRUE, message=FALSE, warning=FALSE}
# Install required package if not already available
if (!require(trend)) install.packages("trend", repos = "https://cloud.r-project.org")



# Create annual time series object
ts_data <- x 
years <- data_all$year

# Fit linear trend
trend_model <- lm(ts_data ~ years)
trend_summary <- summary(trend_model)
fitted_trend <- trend_model$fitted.values

# Plot original series with trend line
plot(ts_data, main = "Original Annual Time Series with Linear Trend",
     ylab = "Value", xlab = "Year", col = "blue", lwd = 2)
lines(years, fitted_trend, col = "red", lwd = 2, lty = 2)
legend("topleft", legend = c("Observed", "Trend line"),
       col = c("blue", "red"), lwd = 2, lty = c(1, 2))

# Detrend series (residuals)
ts_detrended <- residuals(trend_model)

# Plot detrended series
plot(ts_detrended, main = "Detrended Time Series (Residuals)",
     ylab = "Residuals", xlab = "Year", col = "darkgreen", lwd = 2)

# Mann–Kendall trend test
mk_result <- mk.test(ts_data)
mk_result

# Automatic interpretation
if (mk_result$p.value < 0.05) {
  trend_direction <- ifelse(mk_result$estimates[1] > 0, "increasing", "decreasing")
  cat("\nInterpretation: There is a statistically significant", trend_direction,
      "trend (p =", round(mk_result$p.value, 4), ").\n")
} else {
  cat("\nInterpretation: No statistically significant trend detected (p =",
      round(mk_result$p.value, 4), ").\n")
}

# Combined visualization
plot(years, ts_data, type = "l", col = "blue", lwd = 2,
     main = "Original, Trend, and Detrended Series",
     xlab = "Year", ylab = "Q (m³/s)")
lines(years, fitted_trend, col = "red", lwd = 2, lty = 2)
lines(years, ts_detrended + mean(ts_data), col = "darkgreen", lwd = 2, lty = 3)
legend("topleft",
       legend = c("Original", "Trend line", "Detrended (shifted)"),
       col = c("blue", "red", "darkgreen"),
       lwd = 2, lty = c(1, 2, 3))

# --- Trend Summary Table ---
slope <- trend_summary$coefficients[2, 1]
p_value <- trend_summary$coefficients[2, 4]
r2 <- trend_summary$r.squared

trend_table <- data.frame(
  Statistic = c("Slope", "R-squared", "p-value (Linear)", "p-value (Mann–Kendall)"),
  Value = c(round(slope, 4), round(r2, 4), round(p_value, 4), round(mk_result$p.value, 4))
)

knitr::kable(trend_table, caption = "Summary of Trend Analysis")
```

## Normality test for residuals

```{r detrend-trendtest-normality, echo=TRUE, message=FALSE, warning=FALSE}
cat("\n\n### Normality Test for Residuals\n")
shapiro_result <- shapiro.test(ts_detrended)
print(shapiro_result)

# Q–Q plot
qqnorm(ts_detrended, main = "Q–Q Plot of Detrended Residuals")
qqline(ts_detrended, col = "red", lwd = 2)

# Normality interpretation
if (shapiro_result$p.value < 0.05) {
  cat("\nInterpretation: Residuals deviate significantly from normality (p =",
      round(shapiro_result$p.value, 4), ").\n")
} else {
  cat("\nInterpretation: Residuals do not significantly deviate from normality (p =",
      round(shapiro_result$p.value, 4), ").\n")
}
```

## Estimate the spectral density function

```{r}

# ---- Compute raw periodogram ----
spec_raw <- spec.pgram(ts_detrended, taper = 0.1, log = "no", detrend = FALSE, plot = FALSE)

# Convert frequency (cycles/year) → period (years)
periods_raw <- 1 / spec_raw$freq
spec_raw_df <- data.frame(period = periods_raw, spec = spec_raw$spec)

# Find main peak
peak_idx <- which.max(spec_raw$spec)
peak_period <- periods_raw[peak_idx]

# ---- Plot raw periodogram (base R) ----
plot(spec_raw_df$period, spec_raw_df$spec, type = "l", col = "darkgreen", lwd = 2,
     main = "Raw Periodogram (x-axis: Period in Years)",
     xlab = "Period (years)", ylab = "Spectral Density", xlim = c(0, max(periods_raw)))
abline(v = peak_period, col = "red", lwd = 2, lty = 2)
text(peak_period, max(spec_raw_df$spec) * 0.9,
     labels = paste0("Peak ≈ ", round(peak_period, 2), " yrs"),
     pos = 4, col = "red")
grid()

# ---- Smoothed spectral density ----
spec_smooth <- spec.pgram(ts_detrended, spans = c(3, 3), taper = 0.1, log = "no",
                          detrend = FALSE, plot = FALSE)
periods_smooth <- 1 / spec_smooth$freq
spec_smooth_df <- data.frame(period = periods_smooth, spec = spec_smooth$spec)

ggplot(spec_smooth_df, aes(x = period, y = spec)) +
  geom_line(color = "#1f77b4", linewidth = 1.2) +
  geom_vline(xintercept = peak_period, color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = peak_period, y = max(spec_smooth_df$spec) * 0.9,
           label = paste0("Dominant cycle ≈ ", round(peak_period, 2), " yrs"),
           color = "red", hjust = -0.1) +
  scale_x_continuous(limits = c(0, max(periods_smooth)),
                     breaks = pretty(periods_smooth, n = 8)) +
  labs(title = "Smoothed Spectral Density (in Period Domain)",
       x = "Period (years)", y = "Spectral Density") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# ---- AR spectral estimate ----
# spec_ar <- spec.ar(ts_detrended, method = "burg", log = "no", plot = FALSE)
# periods_ar <- 1 / spec_ar$freq
# plot(periods_ar, spec_ar$spec, type = "l", col = "purple", lwd = 2,
#      main = "Parametric AR Spectral Estimate",
#      xlab = "Period (years)", ylab = "Spectral Density", xlim = c(0, max(periods_ar)))
# grid()

# ---- Summary ----
cat("Dominant period detected ≈", round(peak_period, 2), "years\n")
cat("Tip: strong peaks suggest recurring multi-year cycles in the data.\n")
```

## Estimate and plot the ACF

```{r}
## ACF and PACF Side-by-Side using ggplot2

# ---- Compute ACF and PACF ----
acf_data <- Acf(ts_detrended, plot = FALSE)
pacf_data <- Pacf(ts_detrended, plot = FALSE)

# Convert to data frames for ggplot
acf_df <- with(acf_data, data.frame(lag = acf_data$lag, acf= acf_data$acf))
pacf_df <- with(pacf_data, data.frame(lag = pacf_data$lag, pacf= pacf_data$acf))

# ---- Create ACF plot ----
p1 <- ggplot(acf_df, aes(x = lag, y = acf)) +
  geom_bar(stat = "identity", fill = "#1f77b4", color = "black", width = 0.1) +
  geom_hline(yintercept = c(0.0), color = "black") +
  geom_hline(yintercept = c(2, -2)/sqrt(length(ts_detrended)), linetype = "dashed", color = "red") +
  labs(title = "ACF of Detrended Annual Series", x = "Lag", y = "ACF") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank()
  )

# ---- Create PACF plot ----
p2 <- ggplot(pacf_df, aes(x = lag, y = pacf)) +
  geom_bar(stat = "identity", fill = "#2ca02c", color = "black", width = 0.1) +
  geom_hline(yintercept = c(0.0), color = "black") +
  geom_hline(yintercept = c(2, -2)/sqrt(length(ts_detrended)), linetype = "dashed", color = "red") +
  labs(title = "PACF of Detrended Annual Series", x = "Lag", y = "PACF") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank()
  )

# ---- Display side by side ----
grid.arrange(p1, p2, ncol = 2)

# ---- Interpretation ----
cat("Interpretation:\n",
    "- ACF: shows correlation of the series with its own lags.\n",
    "- PACF: shows correlation after removing effects of earlier lags.\n",
    "- Dotted red lines = approximate 95% confidence bounds.\n")

```

# **Time series simulation and forecasting**

## Fit Box–Jenkins (ARIMA) Model to Detrended Series

```{r}
fit_arima <- auto.arima(ts_detrended)
summary(fit_arima)

# Observed vs fitted
years <- data_all$year
fitted_df <- data.frame(
  Year = years,
  Observed = as.numeric(ts_detrended),
  Fitted = as.numeric(fitted(fit_arima))
)

ggplot(fitted_df, aes(x = Year)) +
  geom_point(aes(y = Observed), color = "steelblue", size = 3) +
  geom_line(aes(y = Fitted), color = "red", linewidth = 1.2) +
  labs(title = "ARIMA Model Fit (Detrended Series)",
       x = "Year", y = "Value") +
  theme_minimal(base_size = 14)
```

## Residual Diagnostics

```{r}
res <- residuals(fit_arima)

autoplot(res) +
  labs(title = "Residuals of ARIMA Model", x = "Year", y = "Residuals") +
  theme_minimal(base_size = 14)

# Statistical tests
shapiro_test <- shapiro.test(res)
runs_test <- runs.test(res)
ljung_test <- Box.test(res, lag = 10, type = "Ljung-Box")

test_results <- data.frame(
  Test = c("Shapiro–Wilk (Normality)", "Runs Test (Randomness)", "Ljung–Box (Autocorrelation)"),
  Statistic = c(round(shapiro_test$statistic, 4),
                round(runs_test$statistic, 4),
                round(ljung_test$statistic, 4)),
  `p-value` = c(round(shapiro_test$p.value, 4),
                round(runs_test$p.value, 4),
                round(ljung_test$p.value, 4))
)

test_results$Significance <- ifelse(test_results$p.value < 0.05, "Significant", "Not Significant")

kable(test_results, caption = "Residual Diagnostic Tests for ARIMA Model") %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(4, color = "white",
              background = ifelse(test_results$Significance == "Significant", "#D73027", "#1A9850"))
```

## Forecasting the Detrended Series

```{r}
forecast_horizon <- 10
fc_detrended <- forecast(fit_arima, h = forecast_horizon)

autoplot(fc_detrended) +
  labs(title = "Forecast of the Detrended Series (Box–Jenkins Model)",
       x = "Year", y = "Detrended Value") +
  theme_minimal(base_size = 14)
```

## Reconstructing Forecasts for the Original Series (with Prediction Intervals)

```{r}
# Future years
future_years <- (max(years) + 1):(max(years) + forecast_horizon)

# Predict trend for future years
trend_forecast <- predict(trend_model, newdata = data.frame(years = future_years))

# Ensure both are plain numeric vectors of equal length
trend_forecast <- as.numeric(trend_forecast)
fc_mean <- as.numeric(fc_detrended$mean)
fc_lower80 <- as.numeric(fc_detrended$lower[,1])
fc_upper80 <- as.numeric(fc_detrended$upper[,1])
fc_lower95 <- as.numeric(fc_detrended$lower[,2])
fc_upper95 <- as.numeric(fc_detrended$upper[,2])

# Combine trend + ARIMA forecast
original_forecast_values <- fc_mean + trend_forecast
original_forecast_lower80 <- fc_lower80 + trend_forecast
original_forecast_upper80 <- fc_upper80 + trend_forecast
original_forecast_lower95 <- fc_lower95 + trend_forecast
original_forecast_upper95 <- fc_upper95 + trend_forecast

# Combine into one data frame
forecast_df <- data.frame(
  Year = c(years, future_years),
  Value = c(x, rep(NA, forecast_horizon)),
  Forecast = c(rep(NA, length(x)), original_forecast_values),
  Lower80 = c(rep(NA, length(x)), original_forecast_lower80),
  Upper80 = c(rep(NA, length(x)), original_forecast_upper80),
  Lower95 = c(rep(NA, length(x)), original_forecast_lower95),
  Upper95 = c(rep(NA, length(x)), original_forecast_upper95)
)
```

## Comparison with the Original Trend (with Legend)

```{r}
# Compute the fitted trend for all years
trend_fitted <- predict(trend_model, newdata = data.frame(years = c(years, future_years)))

# Prepare data for ggplot
trend_df <- data.frame(Year = c(years, future_years), Trend = trend_fitted)
historic_df <- data.frame(Year = years, Value = x)

# Comparison plot with legend
ggplot() +
  geom_line(data = historic_df, aes(x = Year, y = Value, color = "Historical Series"), linewidth = 1.2) +
  geom_line(data = trend_df, aes(x = Year, y = Trend, color = "Fitted Linear Trend"), linewidth = 1.2, linetype = "dotted") +
  geom_line(data = forecast_df, aes(x = Year, y = Forecast, color = "ARIMA Forecast"), linewidth = 1.2) +
  geom_ribbon(data = forecast_df, aes(x = Year, ymin = Lower95, ymax = Upper95, fill = "95% Prediction Interval"), alpha = 0.2) +
  geom_ribbon(data = forecast_df, aes(x = Year, ymin = Lower80, ymax = Upper80, fill = "80% Prediction Interval"), alpha = 0.3) +
  geom_vline(xintercept = max(years) + 0.5, linetype = "dashed", color = "black") +
  scale_color_manual(
    name = "Lines",
    values = c("Historical Series" = "steelblue",
               "Fitted Linear Trend" = "darkgreen",
               "ARIMA Forecast" = "darkred")
  ) +
  scale_fill_manual(
    name = "Prediction Intervals",
    values = c("80% Prediction Interval" = "orange", "95% Prediction Interval" = "gray60")
  ) +
  labs(
    title = "Original Series, Fitted Trend, and Forecast Comparison",
    x = "Year", y = "Q (m³/s)",
    caption = "Dotted line: Fitted trend | Shaded areas: 80% and 95% prediction intervals"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom",
        legend.box = "vertical",
        legend.title = element_text(face = "bold"))
```
