---
title: "Time series analysis: daily mean streamflows"
output: html_document
---

# **Call libraries**

```{r callLib}
library(readr)
library(dplyr)
library(moments)   # provides skewness() and kurtosis()
library(reshape2)
library(forecast)
library(gridExtra)
library(knitr)
library(ggplot2)
library(tseries)
library(nortest)
library(kableExtra)
library(randtests)
library(trend)
library(xts)
library(patchwork)  
```

# **Read data**

## Read zip file names

```{r rzip}
zip_files <- list.files("/home/alejandro/Documents/unal/lectures/stochasticMethodsInWR/scripts/data/ElBancoMagMeanDailyDisch", pattern = "\\.zip$", full.names = TRUE)
```

## Read one CSV per zip

```{r rcsv}
data_list <- lapply(zip_files, function(zf) {
  csv_name <- unzip(zf, list = TRUE)$Name[1]   # take the first CSV in zip
  read_csv(unz(zf, csv_name))
})
```

## Combine into one dataframe

```{r cindf}
data_all <- bind_rows(data_list)
```

## Covert a char date column into date

```{r formatdata}
data_all$date <- as.Date(data_all$Fecha, format = "%d/%m/%Y")
```

# **Time series preprocessing**
## Display first rows
```{r displaydata}
head(data_all)
X = data_all$Valor
```

## Show the data structure
```{r datastr}
str(data_all)
```

## Compute statistics summary
```{r computestats}
summary_stats <- data.frame(
  Mean = mean(X, na.rm = TRUE),
  SD = sd(X, na.rm = TRUE),
  Min = min(X, na.rm = TRUE),
  Max = max(X, na.rm = TRUE),
  Median = median(X, na.rm = TRUE),
  Skewness = skewness(X, na.rm = TRUE),
  Kurtosis = kurtosis(X, na.rm = TRUE),
  N = length(X)
)
```

## Print stats
```{r printstats, echo=FALSE}
# Transpose the summary data frame
summary_vertical <- data.frame(
  Statistic = names(summary_stats),
  Value = as.numeric(summary_stats[1, ])
)
kable(summary_vertical, digits = 2, caption = "Summary Statistics of the Time Series")
```

## Plot time series
```{r plotts}
tit <- paste(unique(data_all$Parametro), " at ", unique(data_all$NombreEstacion), sep="") 
ggplot(data_all, aes(x = date, y = Valor)) +
  geom_line(color = "blue", linewidth = 0.5) +
  labs(title = tit, x = "Year", y = "Q (m^3 /s)") +
  theme_minimal()
```

## Plot histogram
```{r plothist}
# Calculate summary statistics
q <- quantile(data_all$Valor, probs = c(0.25, 0.5, 0.75))  # Q1, Median, Q3

# Compute density-based histogram data
h <- hist(X, plot = FALSE)
dens <- h$counts / sum(h$counts) / diff(h$breaks[1:2])  # convert counts to density
ymax <- max(dens)  # max histogram density

ggplot(data_all, aes(x = Valor)) +
  geom_histogram(aes(y = ..density..), bins = 15,
                 fill = "skyblue", color = "white", alpha = 0.6) +
  
  # Density overlay
  geom_density(color = "darkblue", size = 1.2, alpha = 0.7) +
  
  # Vertical lines for mean & quartiles
  geom_vline(xintercept = summary_stats$Mean, color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = q[1], color = "darkgreen", linetype = "dotted", size = 1) +
  geom_vline(xintercept = q[2], color = "blue", linetype = "solid", size = 1) +
  geom_vline(xintercept = q[3], color = "darkgreen", linetype = "dotted", size = 1) +
  
  # Set limits to Y axis
  coord_cartesian(ylim = c(0, ymax)) +
  
  # Labels for stats
  annotate("text", x = summary_stats$Mean, y = 0.02, label = "Mean", color = "red", angle = 90, vjust = -0.5) +
  annotate("text", x = q[2], y = 0.02, label = "Median", color = "blue", angle = 90, vjust = -0.5) +
  
  labs(
    title = "Streamflow Histogram + Density with Mean & Quartiles",
    x = "Q (m³/s)",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)
```

## Q–Q Plot of Time Series
```{r qqplot, echo=TRUE, fig.width=6, fig.height=5}
ggplot(data_all, aes(sample = Valor)) +
  stat_qq(color = "steelblue", size = 2) +
  stat_qq_line(color = "red", linetype = "dashed") +
  labs(title = "Normal Q-Q Plot",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

## Normality Summary of the Time Series
```{r normality-summary, message=FALSE, warning=FALSE}
# Normality tests
#sw <- shapiro.test(X)
ad <- ad.test(X)
jb <- jarque.bera.test(X)

# Combine results
results <- data.frame(
  Statistic = c("Mean", "Std. Deviation", "Skewness", "Kurtosis",
                 "Anderson-Darling (p)", "Jarque-Bera (p)"),
  Value = c(
    summary_stats$Mean,
    summary_stats$SD,
    summary_stats$Skewness,
    summary_stats$Kurtosis,
    ad$p.value,
    jb$p.value
  ),
  Interpretation = c(
    "", "", 
    ifelse(abs(summary_stats$Skewness) < 0.5, "≈ Symmetric", 
           ifelse(summary_stats$Skewness > 0, "Right-skewed", "Left-skewed")),
    ifelse(abs(summary_stats$Kurtosis - 3) < 0.5, "≈ Normal tails", 
           ifelse(summary_stats$Kurtosis > 3, "Leptokurtic (heavy tails)", "Platykurtic (light tails)")),
    ifelse(ad$p.value > 0.05, "Normal", "Not normal"),
    ifelse(jb$p.value > 0.05, "Normal", "Not normal")
  )
)

kable(results, digits = 4, caption = "Summary statistics and normality tests for the time series")
```

## Normalization of the Time Series
```{r normalize-ts}
# Example: standardize time series
Xn <- scale(X)
data_all$ValorN <- Xn

# Check new mean and sd
mean(Xn, na.rm = TRUE)
sd(Xn, na.rm = TRUE)

```

## Comparison of Original and Normalized Time Series
```{r normalized-plot, message=FALSE, warning=FALSE}

# Prepare data
df <- data.frame(
  time = data_all$date,
  Original = as.numeric(X),
  Normalized = as.numeric(Xn)
)

# Plot with dual y-axes
ggplot(df, aes(x = time)) +
  geom_line(aes(y = Original, color = "Original"), size = 1) +
  geom_line(aes(y = Normalized * sd(X) + mean(X), color = "Normalized"), size = 1) +
  scale_y_continuous(
    name = "Original values",
    sec.axis = sec_axis(~ (. - mean(X)) / sd(X),
                        name = "Normalized (Z-score)")
  ) +
  scale_color_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Original and Normalized Time Series",
    x = "Year",
    color = "Series"
  ) +
  theme_minimal() +
  theme(
    axis.title.y.left = element_text(color = "steelblue"),
    axis.title.y.right = element_text(color = "tomato"),
    legend.position = "top"
  )
```

## Q–Q Plot of Normalized Time Series
```{r qqplot_normalized, echo=TRUE, fig.width=6, fig.height=5}
ggplot(data_all, aes(sample = ValorN)) +
  stat_qq(color = "steelblue", size = 2) +
  stat_qq_line(color = "red", linetype = "dashed") +
  labs(title = "Normal Q-Q Plot",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

# Time series analysis
## Randomness Tests
```{r randomness-tests, message=FALSE, warning=FALSE}
# Apply tests
runs <- runs.test(X)
tp   <- turning.point.test(X)
lb   <- Box.test(X, lag = 20, type = "Ljung-Box")

# Combine results
results <- data.frame(
  Test = c("Runs Test", "Turning Point Test", "Ljung-Box (lag=20)"),
  Statistic = c(runs$statistic, tp$statistic, lb$statistic),
  p_value = c(runs$p.value, tp$p.value, lb$p.value),
  Decision = ifelse(c(runs$p.value, tp$p.value, lb$p.value) > 0.05,
                    "Random (fail to reject H₀)",
                    "Not random (reject H₀)")
)

knitr::kable(results, digits = 4, caption = "Tests for randomness in the time series")
```

## Trend Analysis for Annual Time Series
```{r ts-annual-trend, message=FALSE, warning=FALSE}
# Convert to time series
#x <- ts(X, start = min(data_all$year), frequency = 1)
x<- xts(X, order.by = data_all$date)

# --- Trend estimation using a moving average ---
trend <- stats::filter(x, rep(1/3, 3), sides = 2)

# Plot
plot(x, col = "steelblue", lwd = 2, main = "Annual Time Series with Trend", ylab = "Q (m³/s)", xlab = "Year")
lines(trend, col = "tomato", lwd = 2)
legend("topleft", legend = c("Original", "3-Year Moving Average"), col = c("steelblue", "tomato"), lwd = 2)
```

## Detrending a Daily Time Series and Testing for Trend"
```{r}
data_all$t <- as.numeric(data_all$date - min(data_all$date)) + 1

# 2. Fit Linear Trend Model
trend_model <- lm(Valor ~ t, data = data_all)
print(summary(trend_model))

# 3. Plot the Original Series and Fitted Trend
ggplot(data_all, aes(x = date, y = Valor)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_line(aes(y = fitted(trend_model)), color = "red", linewidth = 1) +
  labs(title = "Daily Time Series with Fitted Linear Trend",
       x = "Date", y = "Value") +
  theme_minimal(base_size = 14)

# 4. Detrend the Series (Remove Linear Component)
data_all$residuals <- residuals(trend_model)
data_all$trend <- fitted(trend_model)
ggplot(data_all, aes(x = date, y = residuals)) +
  geom_line(color = "darkgreen") +
  labs(title = "Detrended Daily Time Series (Residuals)",
       x = "Date", y = "Residuals") +
  theme_minimal(base_size = 14)

# 5. Test for Normality of Residuals
cat("\n--- Anderson-Darling Normality Test ---\n")
print(ad.test(data_all$residuals))

# 6. Q-Q Plot of Residuals
qqnorm(data_all$residuals, main = "Q-Q Plot of Detrended Series Residuals")
qqline(data_all$residuals, col = "red", lwd = 2)

# 7. Test for Randomness of Residuals
cat("\n--- Runs Test for Randomness ---\n")
print(runs.test(data_all$residuals))
```
## Estimate the Spectral Density Function for the Detrended Daily Time Series
After removing the linear trend from the daily time series, we can analyze the frequency components of the residuals (detrended series) using the spectral density function. This helps identify dominant periodicities (e.g., seasonal cycles) present in the data.

```{r spectral-density, echo=TRUE, message=FALSE, warning=FALSE}
# Compute the spectral density of the detrended series
spectrum(data_all$residuals,
         main = "Spectral Density of Detrended Daily Time Series",
         xlab = "Frequency",
         ylab = "Spectral Density",
         col = "blue",
         lwd = 2)

# Optionally, use a smoothed periodogram (useful for noisy daily data)
spec_res <- spectrum(data_all$residuals, spans = c(3, 3), plot = FALSE)

# Plot smoothed spectrum
plot(spec_res$freq, spec_res$spec, type = "l", col = "darkred", lwd = 2,
     xlab = "Frequency", ylab = "Spectral Density (Smoothed)",
     main = "Smoothed Spectral Density of Detrended Series")

# Dominant frequency and period
dominant_freq <- spec_res$freq[which.max(spec_res$spec)]
dominant_period_days <- 1 / dominant_freq

cat("Dominant frequency:", round(dominant_freq, 5), "cycles/day\n")
cat("Approximate period:", round(dominant_period_days, 1), "days\n")

```
## ACF, PACF, and Autocorrelation Tests for the Detrended Daily Time Series

```{r acf_pacf_ljung, fig.width=10, fig.height=5, message=FALSE, warning=FALSE}

# --- Ensure detrended data exist ---
if (!exists("data_all") || !"residuals" %in% names(data_all)) {
  stop("Please define a data.frame 'df' with a column 'residuals' (detrended series).")
}

# --- Prepare the residuals ---
res <- na.omit(as.numeric(data_all$residuals))
n <- length(res)
if (n < 4) stop("Not enough data points for ACF/PACF analysis.")

# --- Compute ACF and PACF ---
acf_res <- acf(res, plot = FALSE, lag.max = 365)
pacf_res <- pacf(res, plot = FALSE, lag.max = 365)

acf_df <- data.frame(lag = acf_res$lag[,1,1], acf = acf_res$acf[,1,1])
pacf_df <- data.frame(lag = seq_along(pacf_res$acf), pacf = pacf_res$acf)

# --- Confidence bands ---
conf <- 2 / sqrt(n)

# --- Create plots ---
p_acf <- ggplot(acf_df, aes(x = lag, y = acf)) +
  geom_bar(stat = "identity", width = 0.1, fill = "#1f77b4", color = "black") +
  geom_hline(yintercept = 0, color = "black") +
  geom_hline(yintercept = c(conf, -conf), linetype = "dashed", color = "red") +
  labs(title = "ACF (detrended series)", x = "Lag (days)", y = "ACF") +
  theme_minimal(base_size = 14)

p_pacf <- ggplot(pacf_df, aes(x = lag, y = pacf)) +
  geom_bar(stat = "identity", width = 0.1, fill = "#2ca02c", color = "black") +
  geom_hline(yintercept = 0, color = "black") +
  geom_hline(yintercept = c(conf, -conf), linetype = "dashed", color = "red") +
  labs(title = "PACF (detrended series)", x = "Lag (days)", y = "PACF") +
  theme_minimal(base_size = 14)

# --- Show side-by-side plots ---
(p_acf | p_pacf) + plot_annotation(title = "ACF and PACF of the Detrended Daily Time Series")

# --- Print significant lags ---
sig_acf <- subset(acf_df, lag > 0 & abs(acf) > conf)
sig_pacf <- subset(pacf_df, abs(pacf) > conf)

cat("\nSample size (n) =", n, "\n")
cat("Approx. 95% confidence bounds = ±", round(conf, 4), "\n\n")

if (nrow(sig_acf) > 0) {
  cat("Significant ACF lags (|ACF| > 2/sqrt(n)):\n")
  print(sig_acf)
} else {
  cat("No significant ACF lags detected.\n")
}

if (nrow(sig_pacf) > 0) {
  cat("\nSignificant PACF lags (|PACF| > 2/sqrt(n)):\n")
  print(sig_pacf)
} else {
  cat("\nNo significant PACF lags detected.\n")
}

# --- Ljung–Box test for randomness at multiple lags ---
lags_to_test <- c(10, 20, 30)
lb_results <- data.frame(
  Lag = lags_to_test,
  Statistic = NA,
  p_value = NA
)

for (i in seq_along(lags_to_test)) {
  test <- Box.test(res, lag = lags_to_test[i], type = "Ljung-Box")
  lb_results$Statistic[i] <- test$statistic
  lb_results$p_value[i] <- test$p.value
}

cat("\nLjung–Box Test Results:\n")
print(lb_results)

# --- Interpretation ---
cat("\nInterpretation:\n")
cat("If p-value < 0.05, there is evidence of autocorrelation up to that lag.\n")
cat("If p-value >= 0.05, residuals appear random (no significant autocorrelation).\n")
```

# **Time series simulation and forecasting**
## Fit Box–Jenkins (ARIMA) Model to Detrended Series

---
title: "Box–Jenkins (ARIMA) Model for Detrended Daily Time Series"
output: html_document
---

```{r, echo=TRUE, message=FALSE, warning=FALSE}

# ---- 1. Fit Box–Jenkins (ARIMA) model ----
fit_arima <- auto.arima(data_all$residuals, seasonal = FALSE)
summary(fit_arima)

# ---- 2. Residual diagnostics ----
checkresiduals(fit_arima)

# ---- 2. Statistical tests ----
res <- residuals(fit_arima)

# Normality (Shapiro-Wilk)
if (length(res) > 5000) {
  cat("Shapiro-Wilk test not performed: sample too large.\n")
} else {
  print(shapiro.test(res))
}

# Randomness (Runs test)
if (is.numeric(res)) {
  print(runs.test(res))
}

# Step 3: Simulate fitted values (on detrended scale)
data_all$sim_detrended <- fitted(fit_arima)

# ----------------------------
# Step 4: Add trend back
# ----------------------------
data_all$simulated <- data_all$sim_detrended + data_all$trend

# ----------------------------
# Step 4: Plot full series and zoomed year
# ----------------------------
# Define zoomed subset (last 365 days)
df_zoom <- tail(data_all, 365)

p1 <- ggplot(data_all, aes(x = date)) +
  geom_line(aes(y = Valor, color = "Original"), linewidth = 1) +
  geom_line(aes(y = simulated, color = "Simulated (ARIMA + Trend)"), linewidth = 1, linetype = "dashed") +
  scale_color_manual(values = c("Original" = "blue", "Simulated (ARIMA + Trend)" = "red")) +
  labs(title = "Full Daily Time Series",
       x = "Date", y = "Value", color = "Series") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

p2 <- ggplot(df_zoom, aes(x = date)) +
  geom_line(aes(y = Valor, color = "Original"), linewidth = 1) +
  geom_line(aes(y = simulated, color = "Simulated (ARIMA + Trend)"), linewidth = 1, linetype = "dashed") +
  scale_color_manual(values = c("Original" = "blue", "Simulated (ARIMA + Trend)" = "red")) +
  labs(title = "Zoomed View: Last Year",
       x = "Date", y = "Value", color = "Series") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

# Combine both plots
p1 | p2

# ---- 4. Forecast next 90 days ----
n_forecast <- 90
fc_detrended <- forecast(fit_arima, h = n_forecast)

# ---- 5. Reconstruct full forecast (add trend back) ----
future_dates <- seq(max(data_all$date) + 1, by = "day", length.out = n_forecast)
future_t <- as.numeric(future_dates)
future_trend <- predict(trend_model, newdata = data.frame(t = future_t))

final_forecast <- data.frame(
  date = future_dates,
  forecast = fc_detrended$mean + future_trend,
  lower = fc_detrended$lower[, 2] + future_trend,
  upper = fc_detrended$upper[, 2] + future_trend
)

# ---- 6. Plot full series with trend + forecast + legend ----
ggplot() +
  geom_line(data = data_all, aes(x = date, y = Valor, color = "Observed"), linewidth = 0.6) +
  geom_line(data = data_all, aes(x = date, y = trend, color = "Trend"), linewidth = 0.8, linetype = "dashed") +
  geom_line(data = final_forecast, aes(x = date, y = forecast, color = "Forecast"), linewidth = 0.9) +
  geom_ribbon(data = final_forecast, aes(x = date, ymin = lower, ymax = upper, fill = "95% CI"), alpha = 0.3) +
  scale_color_manual(name = "Series", values = c("Observed" = "gray40", "Trend" = "red", "Forecast" = "blue")) +
  scale_fill_manual(name = "Interval", values = c("95% CI" = "lightblue")) +
  labs(title = "Box–Jenkins (ARIMA) Forecast with Linear Trend (Daily Series)",
       x = "Date", y = "Value") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")

# ---- 7. Zoomed-in plot with legend ----
zoom_days <- 180
data_zoom <- data_all %>% filter(date >= max(date) - zoom_days)

ggplot() +
  geom_line(data = data_zoom, aes(x = date, y = Valor, color = "Observed"), linewidth = 0.6) +
  geom_line(data = data_zoom, aes(x = date, y = trend, color = "Trend"), linewidth = 0.8, linetype = "dashed") +
  geom_line(data = final_forecast, aes(x = date, y = forecast, color = "Forecast"), linewidth = 0.9) +
  geom_ribbon(data = final_forecast, aes(x = date, ymin = lower, ymax = upper, fill = "95% CI"), alpha = 0.3) +
  scale_color_manual(name = "Series", values = c("Observed" = "gray40", "Trend" = "red", "Forecast" = "blue")) +
  scale_fill_manual(name = "Interval", values = c("95% CI" = "lightblue")) +
  labs(title = "Zoomed Forecast with Trend",
       x = "Date", y = "Value") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

